---
title: "Interactive Data Viz"
author: "Your Name Here"
date: "2023-11-27"
output: html_document
editor_options: 
  chunk_output_type: console
---
# Interactive Plots
## Plotly

### Scatterplot with Plotly

This code creates a simple scatterplot using the plotly library with the penguins dataset. The scatterplot shows the relationship between bill length and bill depth colored by species. Here's a breakdown of the code:

* plot_ly(): This function initializes the creation of a Plotly graph.
* data = penguins: Specifies the dataset to be used, in this case, the Iris dataset.
* x = ~bill_length_mm, y = ~bill_depth_mm, color = ~species: Defines the x and y variables for the plot and how the points will be colored.
* type = "scatter", mode = "markers": Indicates that the type of plot is a scatter plot with markers(points).
* layout(): This function sets the layout of the plot, including title, axis labels, and background color.
The resulting fig object is an interactive scatterplot where you can hover over points to see details, pan, zoom, and perform other interactions.


```{r}
library(plotly)
library(palmerpenguins)
data("penguins")

fig <- plot_ly(data = penguins, x = ~bill_length_mm, y = ~bill_depth_mm, color = ~species, 
               type = "scatter", mode = "markers")%>%
  layout(title="A Plotly Figure", legend=list(title=list(text='species')),
         plot_bgcolor='#e5ecf6', 
         xaxis = list( 
           zerolinecolor = '#ffff', 
           zerolinewidth = 2, 
           gridcolor = 'ffff'), 
         yaxis = list( 
           zerolinecolor = '#ffff', 
           zerolinewidth = 2, 
           gridcolor = 'ffff'))
fig
```

### More Basic Scatterplot with Plotly

This code creates another scatterplot using the Iris dataset, but this time, it shows the relationship between bill length and flipper length. Here's a breakdown:

* The data variables x = ~bill_length_mm, y = ~flipper_length_mm specify the x and y axes.
* The plot type is scatter with markers.
This scatterplot, like the previous one, is interactive and allows for exploration of data points.

```{r}
library(plotly)

fig2 <- plot_ly(data = penguins, x = ~bill_length_mm, y = ~flipper_length_mm, type = "scatter", mode = "markers")

fig2
```

###Plotting Markers and Lines

This example is about creating a line plot with multiple traces. It uses randomly generated data. Here's what's happening:

* Three random vectors (trace_0, trace_1, and trace_2) are created with 100 data points each.
* The data data frame is created, containing these three vectors and a sequence of numbers x from 1 to 100.
* plot_ly() initializes a plot object using this data frame.
* Three traces are added to the plot using add_trace(). Each trace corresponds to one of the random vectors and is designated by a name ('trace 0', 'trace 1', 'trace 2').

This results in a plot with three traces, where you can visualize multiple lines and markers.

```{r}

trace_0 <- rnorm(100, mean = 5)
trace_1 <- rnorm(100, mean = 0)
trace_2 <- rnorm(100, mean = -5)
x <- c(1:100)

data <- data.frame(x, trace_0, trace_1, trace_2)

fig <- plot_ly(data, x = ~x)
fig <- fig %>% add_trace(y = ~trace_0, name = 'trace 0',mode = 'lines')
fig <- fig %>% add_trace(y = ~trace_1, name = 'trace 1', mode = 'lines+markers')
fig <- fig %>% add_trace(y = ~trace_2, name = 'trace 2', mode = 'markers')

fig
```

## Combining ggplot2 with Plotly

Here, the code demonstrates how to convert a ggplot2 plot into an interactive Plotly plot. The plot is created using the babynames dataset and represents the popularity of names of The Beatles members. Here's the explanation:

* A ggplot object called beatles_plot is created using ggplot2.
* The ggplotly() function is used to convert the ggplot plot to a Plotly interactive plot.

```{r}
library(tidyverse)
library(mdsr)
library(babynames)
Beatles <- babynames %>%
  filter(name %in% c("John", "Paul", "George", "Ringo") & sex == "M") %>%
  mutate(name = factor(name, levels = c("John", "George", "Paul", "Ringo")))
beatles_plot <- ggplot(data = Beatles, aes(x = year, y = n)) +
  geom_line(aes(color = name), size = 2)
beatles_plot

ggplotly(beatles_plot)
```

This shows that you can enhance static ggplot2 plots with interactivity by using ggplotly().

## Interactive Tables with Datatable

In this section, an interactive table is generated using the DT package. The datatable() function is used to create the table from the Beatles dataset. Here's how it works:

* DT::datatable() initializes an interactive table using the Beatles dataset.
* The options argument can be used to set various table options, like the number of rows to display per page.

The result is an interactive table that allows you to sort, search, and explore the data efficiently.

```{r}
library(DT)
datatable(Beatles, options = list(pageLength = 10))
```

### Dygraph
This code generates an interactive time series plot showing the popularity of Beatles names over time. You can adjust the date range and explore the data dynamically. Dygraphs is a powerful tool for visualizing time-dependent information and can be customized further to suit your needs.

* We filter the Beatles dataset for male names.
* We select the relevant columns, such as year, name, and prop (popularity).
* We use pivot_wider to reformat the data.
* dygraph(main = "Popularity of Beatles names over time") initializes a Dygraph plot and sets the title.
* dyRangeSelector(dateWindow = c("1900", "1980")) provides a range selector to focus on specific years.

```{r}
library(dygraphs)
#Dygraphs
Beatles %>% 
  filter(sex == "M") %>% 
  select(year, name, prop) %>%
  pivot_wider(names_from = name, values_from = prop) %>%
  dygraph(main = "Popularity of Beatles names over time") %>% 
  dyRangeSelector(dateWindow = c("1900", "1980"))
```

## In class exercises

Make an interactive plot that shows pass rate by month with inspection type represented by color using the inspection data from last week. Repeat this for a dygraph.

```{r}


```