---
title: "Strings and Dates"
author: "Your Name Here"
date: "2023-10-23"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Strings and Dates

## Introduction to stringr
stringr is a powerful R package in the tidyverse for working with strings. It provides a set of functions that make it easier to manipulate, analyze, and extract information from character strings. In this tutorial, you'll learn how to use stringr for common string operations .Please see the [cheatsheet](https://github.com/rstudio/cheatsheets/blob/main/strings.pdf) for more details.

### Installing and Loading the stringr Package
Before you can use stringr, you need to install and load the package. Open R or RStudio and run the following commands:

```{r string}
# Install stringr (only need to do this once)
#install.packages("stringr")

# Load stringr
library(stringr)

# Also in tidyverse
library(tidyverse)
```

### Basic String Operations
Let's start with some basic string operations.

1. str_length(): Finding the Length of a String
You can use str_length() to find the length (the number of characters) in a string:

```{r length}
# Example
my_string <- "Hello, World!"
str_length(my_string)  # Output: 13
```

2. str_to_lower() and str_to_upper(): Changing Case
You can use str_to_lower() and str_to_upper() to change the case of a string:

```{r cap}
str_to_lower(my_string)  # Output: "hello, world!"
str_to_upper(my_string)  # Output: "HELLO, WORLD!"
```

### Pattern Matching
stringr provides functions to perform pattern matching and extraction.

3. str_detect(): Detecting Patterns
Use str_detect() to check if a string contains a specific pattern:

```{r detect}
# Example
my_string <- "The quick brown fox jumps over the lazy dog."
str_detect(my_string, "fox")  # Output: TRUE
``` 

4. str_replace(): Replacing Patterns
You can use str_replace() to replace a pattern with another string:

```{r replace}
# Example
my_string <- "The quick brown fox jumps over the lazy dog."
str_replace(my_string, "fox", "cat")  # Output: "The quick brown cat jumps over the lazy dog."
``` 

5. str_extract(): Extracting Patterns
str_extract() extracts the first occurrence of a pattern from a string:

```{r extract}
# Example
my_string <- "Email me at john@example.com or call at 555-123-4567."
str_extract(my_string, "\\d{3}-\\d{3}-\\d{4}")  # Output: "555-123-4567"
```

6. str_split(): Splitting Strings
str_split() splits a string into a character vector using a specified delimiter:

```{r split}
# Example
my_string <- "apple,banana,cherry"
str_split(my_string, ",")  # Output: a character vector ["apple", "banana", "cherry"]
``` 

String Cleaning
7. str_trim(): Trimming Whitespace
Use str_trim() to remove leading and trailing whitespace from a string:

```{r trim}
# Example
my_string <- "  Trim me!   "
str_trim(my_string)  # Output: "Trim me!"
```

8. str_squish(): Removing Extra Whitespace
str_squish() removes extra whitespace within a string:

```{r squish}
# Example
my_string <- "Hello   world    from    R!"
str_squish(my_string)  # Output: "Hello world from R!"
``` 

String Manipulation
9. str_c(): Combining Strings
You can use str_c() to concatenate strings together:

```{r conc}
# Example
first_name <- "John"
last_name <- "Doe"
full_name <- str_c(first_name, " ", last_name)
``` 

10. str_sub(): Substring Extraction
str_sub() extracts substrings from a string:

```{r sub}
# Example
my_string <- "Extract this part"
str_sub(my_string, start = 9, end = 13)  # Output: "this"
``` 

### Summary
stringr is a versatile package for working with strings in R. It provides functions for string manipulation, pattern matching, and cleaning. These functions make it easier to perform various text-related tasks when working with data.

To get more familiar with string manipulation in R, try out these functions with different strings and patterns. The more you practice, the more proficient you'll become in handling strings in R.

## On Your Own..

Using the `sentences` dataset (which is a vector may have to change to a data.frame), practice the following questions.

1. Calculate and report the average character length of sentences in the "sentences" dataset. Which sentence is the shortest, and which is the longest?

2. Find all the sentences in the "sentences" dataset that use a comma. How many such sentences are there?

```{r}
data.frame(sentences) %>% 
  filter(str_detect(sentences, ","))
##
data.frame(sentences) %>%
  mutate(comma = str_detect(sentences, ",")) %>% 
  summarise(prop = sum(comma)/n())

```


3. Count how many occurrence of the letter "t" is in each sentence. Which sentence has the most? How many have 0?

```{r}
data.frame(sentences) %>% 
  mutate(ts = str_to_lower(sentences)) %>% 
  mutate(ts = str_count(ts, "t")) %>% 
  summarise(max = max(ts),
            zeros = sum(ts == 0))

## stupid way
data.frame(sentences) %>% 
  mutate(ts = str_to_lower(sentences)) %>%
  mutate(ts = str_extract_all(sentences, "t")) %>% 
  mutate(count = map(ts, ~as.numeric(length(.)))) %>% 
  summarise(max = max(unlist(count)),
            zeros = sum(unlist(count) == 0))

```


4. In the "sentences" dataset, extract and list all the sentences that contain the word "and" How many sentences contain this word?

```{r}

data.frame(sentences) %>% 
  filter(str_detect(sentences, "and"))

```


5. Replace all occurrences of the word "good" with "excellent" in the "sentences" dataset. How many replacements were made, and what does the dataset look like after these replacements?

```{r}
data.frame(sentences) %>% 
  mutate(change = str_detect(sentences, "good")) %>% 
  mutate(new_sent = ifelse(change == "TRUE", 
                           str_replace_all(sentences, "good", "excellent"),
                           NA)) %>% 
  filter(!is.na(new_sent))

data.frame(sentences) %>% 
  filter(str_detect(sentences, "good")) %>% 
  mutate(new_sent = str_replace_all(sentences, "good", "excellent"))
```


##Introduction to lubridate
lubridate is a popular R package that makes it easy to work with date-times. It provides a wide range of functions and methods to parse, manipulate, and format date-time objects. This tutorial will cover some of the fundamental aspects of using lubridate for working with date and time data in R. Please see the [cheatsheet](https://rawgit.com/rstudio/cheatsheets/main/lubridate.pdf) for more details.

### Installing and Loading lubridate
To use the lubridate package, you need to install it first. Open R or RStudio and run the following commands:

```{r lubridate}
# Install lubridate (only need to do this once)
#install.packages("lubridate")

# Load lubridate
library(lubridate)
```

Now, you're ready to start working with date-time objects.

### Creating Date-Time Objects
You can create date-time objects in various ways:

1. Using ymd() for Date
The ymd() function is used to create a date object from year, month, and day components.

```{r lab}
my_date <- ymd("2023-09-15")
```

2. Using mdy() for Date
If you have a date in the month-day-year format, you can use the mdy() function.

```{r lab2}
my_date <- mdy("09-15-2023")
```

3. Using ymd_hms() for Date-Time
To create a date-time object with hours, minutes, and seconds, use the ymd_hms() function.

```{r lab3}
my_datetime <- ymd_hms("2023-09-15 14:30:45")
```

### Extracting Components
You can extract various components of date-time objects:

4. Extracting Year, Month, Day, etc.
```{r extract-date}
year(my_date)  # Extract year
month(my_date)  # Extract month
day(my_date)  # Extract day
wday(my_date)  # Extract day of the week (Sunday: 1, Monday: 2, etc.)
```

### Arithmetic with Date-Times
You can perform arithmetic operations with date-time objects:

5. Adding Days
```{r adding}
new_date <- my_date + days(7)  # Add 7 days to my_date
```

6. Finding Time Difference
```{r diff}
diff <- my_date - my_datetime  # Find the time difference between my_date and my_datetime


## can use difftime() too
time1 <- ymd_hms("2023-09-15 08:30:00")
time2 <- ymd_hms("2023-09-15 12:45:30")

# Calculate the time difference in seconds
time_diff_seconds <- as.numeric(difftime(time2, time1, units = "secs"))

# Calculate the time difference in hours
time_diff_hours <- as.numeric(difftime(time2, time1, units = "hours"))
```

7. Finding Intervals
lubridate provides functions to find intervals between two date-time objects. For example, you can use interval() to create a time interval between two date-times:

```{r int}
start_time <- ymd_hms("2023-09-15 08:30:00")
end_time <- ymd_hms("2023-09-15 12:45:30")

# Create an interval
time_interval <- interval(start_time, end_time)

```


### Formatting Dates
You can format date-time objects for display:

8. Using format()
```{r format-labs}
formatted_date <- format(my_date, format = "%A, %B %d, %Y")
```

### Parsing Dates
You can parse character strings into date-time objects:

9. Using parse_date_time()
```{r parse}
date_string <- "Sep 15, 2023 14:30:45"
parsed_datetime <- parse_date_time(date_string, orders = "mdy HMS")
```

### Summary
lubridate is a powerful package in R for handling date-time data. It simplifies the creation, extraction, arithmetic, formatting, and parsing of date-time objects. This tutorial covered some of the basic operations you can perform with lubridate. As you explore date-time data, you'll find that lubridate is a valuable tool for your R programming tasks.

### On your Own..
 Using the `flights` dataset answer the following questions.
 
 1. Create a new column in the "flights" dataset that displays the departure time in the "hh:mm:ss" format (e.g., "08:45:00")
 
```{r}
library(nycflights13)
data("flights")

flights %>% 
  mutate(dep_time = format(parse_date_time(dep_time, orders = "HM")),
         format = "%H:%M:%S")%>% 
  select(dep_time)

flights %>% 
  mutate(dep_time = str_c(year, month, day, dep_time)) %>% 
  mutate(dep_time = ymd_hm(dep_time)) %>%
  select(dep_time)

flights %>% 
  mutate(dep_time = format(parse_date_time(dep_time, orders = "HM")),
         format = "%H:%M:%S")%>% 
  mutate(dep_time = str_c(hour(dep_time),":", minute(dep_time))) %>%
  mutate(dep_time = str_c(year, "-", month, "-", day, " ", dep_time)) %>% 
  mutate(dep_time = ymd_hm(dep_time)) %>% 
    select(dep_time)
```
 
 2.Extract the year, month, and day from the "dep_time" column in the "flights" dataset and create separate columns for each.
 
 3. Calculate the average departure delay for flights that occurred on Fridays. How does this compare to the average departure delay on Mondays?
 
```{r}
flights %>% 
  mutate(dep_time = ymd_hm(str_c(year, "-", month, "-", day, " ",
                          hour(parse_date_time(dep_time, orders = "HM")),
                          ":",
                          minute(parse_date_time(dep_time, orders = "HM")))),
         sched_dep_time = ymd_hm(str_c(year, "-", month, "-", day, " ",
                          hour(parse_date_time(sched_dep_time, orders = "HM")),
                          ":",
                          minute(parse_date_time(sched_dep_time, orders = "HM"))))) %>% 
  select(dep_time, sched_dep_time) %>% 
  mutate(diff = difftime(dep_time, sched_dep_time, units = "mins")) %>% 
  mutate(day_of_week = wday(dep_time, label = TRUE)) %>% 
  group_by(day_of_week) %>% 
  summarise(ave = mean(diff))
  

### different way not using time

flights %>% 
  mutate(date = ymd(str_c(year,"-", month, "-", day))) %>% 
  mutate(day_of_week = wday(date, label = TRUE)) %>% 
  group_by(day_of_week) %>% 
  summarise(ave = mean(dep_delay, na.rm = TRUE))

### interesting slightly different results
```

 


## Inside the Tidyverse

Using these packages along with Tidyverse functions flow together inside the pipe.

Pulling out the first word using `\\w+` and more can be found [here](https://stringr.tidyverse.org/articles/regular-expressions.html). 

```{r string-pipe}
# Sample dataset
data <- tibble(name = c("John Doe", "Jane Smith", "Bob Johnson"),
               score = c(85, 92, 78))

# Let's say we want to extract first names from the 'name' column
data <- data %>%
  mutate(first_name = str_extract(name, "\\w+"))

# Now, filter the data for scores greater than 80
high_scorers <- data %>%
  filter(score > 80)

# Print the result
print(high_scorers)
```

Similar ideas can be applied to dates like using `floor_date()`

```{r}
# Sample dataset with dates and values
data <- tibble(date = c("2023-05-01", "2023-05-20", "2023-05-31"),
               value = c(45, 62, 73))

# Convert the 'date' column to a Date class
data <- data %>%
  mutate(date = ymd(date))

# Floor the dates to the beginning of the month
data <- data %>%
  mutate(week = floor_date(date, unit = "week"))

# Filter the data for dates in May 2023
may_data <- data %>%
  filter(week == ymd("2023-05-28"))

# Print the result
print(may_data)
```


