---
title: "Wrangle 2"
date: "2023-09-11"
output: html_document
---
## Data Transformation
### select()
> When working with select think about selecting COLUMNS

There are different ways to use `select()`

1. Choose specific column name from the data frame
```
    # Select columns "age" and "income"
    selected_data <- select(data, age, income)
```
2. Select Columns by Name Patterns:
You can use special helper functions like `starts_with()`, `ends_with()`, `contains()`, `matches()`, and `everything()` to select columns based on their names.

```
    # Select columns that start with "var"
    selected_vars <- select(data, starts_with("var"))
    
    # Select columns that contain "total" in their names
    total_columns <- select(data, contains("total"))
```
3. Exclude Columns:
To exclude specific columns, you can use the `-` (minus) sign before the column name.
```
    # Exclude columns "gender" and "zipcode"
    filtered_data <- select(data, -gender, -zipcode)
```
4. Select Columns by Index and Range:
Using the : operator, you can select columns within a range of indices based on their index positions.
```
    # Select columns 4 to 6
    range_columns <- select(data, 4:6)
```
#### Practice with `CIACountries`

We will load the `library(mdsr)` (might have to install) which comes from *Modern Data Science with R* book and houses the data set `CIACountries`

**CIA Countries Description**: The CIA Factbook has geographic, demographic, and economic data on a country-by-country basis. In the description of the variables, the 4-digit number indicates the code used to specify that variable on the data and documentation website.


```{r cia, message=FALSE, results='hide'}
library(tidyverse)
library(mdsr)
## glimpse function in the tidyverse to explore the dataset
CIACountries %>% 
  glimpse()

#Subset of columns
select(CIACountries, country, pop, gdp)

#I can also do it like this
CIACountries %>% 
  select(country, pop, gdp)

```

Explore on your own...

```{r select-ex, results='hide'}
CIACountries %>% 
  select()
```



### filter()
> When filtering think about filtering ROWS

Filter uses Boolean logic so you must enter conditions that are either `TRUE` of a variable(s) or `FALSE`.

Ways to filter..

1. Simple Conditions - greater than (`>`), less than (`<`), or equal to (`==`)
```
    # Select rows where "age" is greater than 30
    filtered_data <- filter(data, age > 30)
    
    # Select rows with "female" in the "gender" column
    female_data <- filter(data, gender == "female")
```
> Must use == when filtering since it is asking about Boolean logic
> Must be exact matches when characters (spelling/uppercase/lowercase/etc.)

2. Multiple Conditions - You can combine conditions using logical operators like `&` (AND) and `|` (OR).

```
    # Select rows where "age" is greater than 30 and "income" is less than 50000
    filtered_data <- filter(data, age > 30 & income < 50000)
```

3. Exclusion - to exclude certain rows, you can use the `!=` operator (not equal to).

```
    # Exclude rows with "zipcode" equal to 12345
    filtered_data <- filter(data, zipcode != 12345)
```
4. Filter rows based on vector of conditions - The `%in%` operator is useful for filtering rows with values in a specified vector.

```
    # Select rows where "gender" is either "male" or "female"
    gender_data <- filter(data, gender %in% c("male", "female"))
```
#### Practice with CIACountries

```{r filter_cia, results='hide'}
#Subset of rows
filter(CIACountries, pop > 1000000000)

#I can also do it like this
CIACountries %>% 
  filter(pop > 1e9)
# using both together
CIACountries %>% 
  select(country, pop, gdp) %>% 
  filter(pop > 1e9)
```

On your own...

Try using filter and select together. 

```{r filter_ex, eval=FALSE}
CIACountries %>% 
  filter() %>% 
  select()
```

### mutate()

There are multiple ways to use the `mutate()` function when creating a new variable depending on what you want to do with the new variable using variables already existing in the dataset.

```{r mutate1, eval=FALSE}
# Add a new variable "bmi" calculated from "weight" and "height"
new_data <- data %>%
  mutate(bmi = weight / (height * height))
# Modify the "income" variable by multiplying it with 1.1
updated_income <- data %>%
  mutate(income = income * 1.1)
```

We can use other functions within mutate to help us make the new variable like `ifelse()`. If you have multiple condition you can use the `case_when()` function and list out your possible options.

Basic syntax is

```
    case_when(
      condition_1 ~ value_1,
      condition_2 ~ value_2,
      ...,
      TRUE ~ default_value
    )
```
* condition_1, condition_2, etc.: These are logical expressions that define conditions to be evaluated. If a condition is TRUE, the corresponding value will be assigned.

* value_1, value_2, etc.: These are the values to be assigned if the corresponding condition is met.

* TRUE ~ default_value: This is optional. If none of the preceding conditions are met, the default_value will be assigned.

```
# Create a new variable "status" based on "age"
new_data <- data %>%
  mutate(status = ifelse(age > 30, "Older", "Younger"))

# Create a new variable "group" based on "age"
new_data <- data %>%
  mutate(group = case_when(
    age < 30 ~ "Young",
    age >= 30 & age <= 50 ~ "Middle-aged",
    age > 50 ~ "Old"
  ))
```


#### Practice with CIACountries

```{r mutate_cia}
#Mutate
#Population Density
my_CIACountries <- mutate(CIACountries, dens = pop/area)
my_CIACountries[1,]

my_CIACountries <- CIACountries %>% 
  mutate(dens = pop/area)
my_CIACountries[1,]
```

On your own..

```{r mutate_ex, eval=FALSE}
CIACountries %>% 
  mutate()
```

## Data Summary

Using `group_by()` and `summarize()` together allows you to efficiently compute summary statistics, aggregations, or any other computations of data based on different groups defined by one or more variables.

### group_by()

The group_by() function is used to group a data frame by one or more variables. This creates a "grouped" data frame where subsequent operations are performed within each group separately. This works best with categorical variables or factor variables.

```
    # Group data by "category"
    grouped_data <- data %>%
      group_by(category)
```

This does not really change the look of the data on its own but returns the dataset with now grouped variables.

### summarise()

The `summarise()` function is used to compute summary statistics or other values for each group. It condenses the grouped data into a single row per group, summarizing the specified variables.

```
    # Compute mean and median of "value" for each group
    summary_results <- grouped_data %>%
      summarise(mean_value = mean(value),
                median_value = median(value))
``` 

Combining `group_by()` and `summarise()` allows you to compute summary statistics or perform calculations on subsets of your data based on grouping variables.

```{r groupby_sum}
# Sample data frame
data <- data.frame(category = c("A", "B", "A", "B", "A"),
                   value = c(10, 20, 15, 25, 30))

# Group by "category" and compute mean and median of "value"
summary_data <- data %>%
  group_by(category) %>%
  summarise(mean_value = mean(value),
            median_value = median(value))
```

#### Practice with CIACountries

```{r summarize_cia, results='hide'}
#What is the average area by high vs low education countries?
my_CIACountries <- CIACountries %>% 
  mutate(high_educ = educ > 4.5) %>% 
  group_by(high_educ) %>% 
  summarise(mn_area = mean(area))

my_CIACountries

```

On your own..

Create a new categorical variable (using `case_when()`) to use `group_by()` and `summarise()`

```{r summarise_ex, eval = FALSE}
CIACountries %>% 
  mutate(new_variable = case_when(
    
  )) %>% 
  group_by(new_variable) %>% 
  summarise()
```


## Data Aggregation 
### arrange() and count()

The `arrange()` function is used to reorder the rows of a data frame based on one or more variables. Depending on the type of variable (integer, character, etc.) depends on how the variables is arranged.

The default order is from largest to smallest (or alphabetical for strings). To change the order from smallest to largest use `desc()`

Unlike other `dplyr` verbs, `arrange()` largely ignores grouping; you need to explicitly mention grouping variables (or use `.by_group = TRUE`) in order to group by them, and functions of variables are evaluated once per data frame, not once per group.

The `count()` function is used to count the occurrences of unique values in a variable. You can also combine these functions to sort the data by a count of occurrences.

```{r arrange, results='hide'}
# Sample data frame
data <- data.frame(category = c("A", "B", "A", "B", "A"),
                   age = c(25, 30, 35, 40, 45))

# Arrange data by "age" in descending order
data %>%
  arrange(desc(age))

# Count the occurrences of each "category"
data %>%
  count(category)

# Arrange data by the count of each "category" in ascending order
data %>%
  count(category) %>%
  arrange(n)
```

### Practice CIA Countries 

```{r arrange_cia}
#Defaults to ascending
CIACountries %>% 
  arrange(gdp) %>% 
  head(5)
#Sort descending
CIACountries %>% 
  arrange(desc(gdp)) %>% 
  head(5)

```

## On your own

Explore CIA Countires dataset, ask questions like "What proportion of countries produce any oil stratified by net users class?"

```{r explore_ex, eval=FALSE}
CIACountries
```

